---
output:

  html_document: default
  pdf_document: default
---


we c---
title: "Assignment 2.1"
output: html_notebook

---

# Description of Demographics
In this section, describe all the demographic variables that you intend to use in your
analysis. In addition to your write-up, it should include relevant numerical measures
(including tables) and graphs.


Name: Sammarieo Brown
ID: 620142596



### Package Management -> Importing the necessary packages that will be used in this project
```{r}

library(tibble)
library(tidyr)
library(haven)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(ggrepel)


```
### Date Pre-processing & Preparation
1. Import dataset (SLC_2007.sav)
2. Rename column headers to be more descriptive.
3. convert the .sav file to a .csv file
```{r}
dataframe <- read_sav("F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.sav")

# rename column names to a more descriptive column name

colnames(dataframe)[1:45] <- c("HH_Num", "Final_Weight", "Water_Bill", "HH_Size_All", "HH_Size_Mem", "Per_Cap_Con_All", "Per_Cap_Con_Mem", "Per_Cap_Pop_Quint", "Type_Dwelling", "Material_Walls", "Num_Rooms", "Type_Toilet", "Toilet_Shared", "Kitchen_Shared", "Own_Dwelling", "Other_Dwelling", "Renter", "Rent_Amt", "Rent_Period", "Rent_Helper", "Pay_Mortgage", "Mortgage_Payment", "Mortgage_Num", "Mortgage_Period", "Pay_Taxes", "Tax_Period", "Water_Source", "Water_Lock", "Water_Meter", "Water_Bill_Latest", "Water_Bill_Months", "Water_Source_Shared", "Water_Source_Dist", "Water_Source_Dist_Code", "Light_Source", "Electric_Bill", "Light_Bill_Months", "Own_Land_Tele", "Own_Cell_Tele", "Land_Tele_Bill", "Cell_Tele_Bill", "Land_Tele_Use", "Cell_Tele_Use", "Garbage_Dispose", "Area_Code")

# convert all the column names to lower case
colnames(dataframe) <- tolower(colnames(dataframe))

# coverting the SLC_2007.sav to csv

converted_df <- write.table(x=dataframe,file="F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.csv", quote = FALSE, sep = ",")

```


Set the seed and take 90% of the random sample



```{r}

set.seed(710)

# import dataset
SLC_2007.Data <- read.csv("F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.csv")


# random sample
n <- nrow(SLC_2007.Data)
sample_size <- round(0.9 * n) # calculate the desired sample size as 90% of the total number of rows

SLC_2007.Sample <- SLC_2007.Data[sample(seq_len(n), size = sample_size, replace = FALSE),] # take a random sample of the data set without replacement


print(colnames(SLC_2007.Sample))
```
```{r}
# create a sub dataset with the demographic variables
SLC_2007.Subset <- select(SLC_2007.Sample,
                          area_code,type_dwelling,
                          type_toilet,toilet_shared,kitchen_shared,own_dwelling,
                          water_source,area_code,hh_size_all,hh_size_mem,
                          per_cap_con_all,water_bill,water_source, water_bill_latest

)
print(colnames(SLC_2007.Subset))

print(head(SLC_2007.Subset))
```
## Data Wrangling
This process of the Data Science lifecycle involves cleaning, transforming and restructuring  the raw data to make it suitable for analysis.



```{r}

# rename all the elements of the rows in our subset to labels.

SLC_2007.Subset <- SLC_2007.Subset %>%
  rename(
    area_code = area_code,
    type_dwelling = type_dwelling,
    type_toilet = type_toilet,
    toilet_shared = toilet_shared,
    kitchen_shared = kitchen_shared,
    own_dwelling = own_dwelling,
    water_source = water_source,

  ) %>%
  mutate(
    area_code = case_when(
      area_code == 1 ~ "KMA",
      area_code == 2 ~ "Other Town",
      area_code == 3 ~ "Rural",
      TRUE ~ as.character(area_code) # keep original value if not matched
    ),
    type_dwelling = case_when(
      type_dwelling == 1 ~ "SEPARATE HOUSE DETACHED",
      type_dwelling == 2 ~ "SEMI-DETACHED HOUSE",
      type_dwelling == 3 ~ "PARTS OF A HOUSE",
      type_dwelling == 4 ~ "APARTMENT BUILDING",
      type_dwelling == 5 ~ "TOWNHOUSE",
      type_dwelling == 6 ~ "IMPROVISED HOUSING UNIT",
      type_dwelling == 7 ~ "PARTS OF COMMERCIAL BUILDING",
      type_dwelling == 8 ~ "OTHER (SPECIFY)",
      TRUE ~ as.character(type_dwelling) # keep original value if not matched
    ),

    type_toilet = case_when(
      type_toilet == 1 ~ "W.C. LINKED TO SEWER",
      type_toilet == 2 ~ "W.C. NOT LINKED",
      type_toilet == 3 ~ "PIT",
      type_toilet == 4 ~ "OTHER",
      type_toilet == 5 ~ "NONE",
      TRUE ~ as.character(type_toilet) # keep original value if not matched
    ),
    toilet_shared = case_when(
      toilet_shared == 1 ~ "EXCLUSIVE USE",
      toilet_shared == 2 ~ "SHARED",
      TRUE ~ as.character(toilet_shared) # keep original value if not matched
    ),
    kitchen_shared = case_when(
      kitchen_shared == 1 ~ "EXCLUSIVE USE",
      kitchen_shared == 2 ~ "SHARED",
      kitchen_shared == 3 ~ "NONE",
      TRUE ~ as.character(kitchen_shared) # keep original value if not matched
    ),
    own_dwelling = case_when(
      own_dwelling == 1 ~ "YES",
      own_dwelling == 2 ~ "NO",
      TRUE ~ as.character(own_dwelling) # keep original value if not matched
    ),
    water_source = case_when(
      water_source == 1 ~ "Indoor tap/pipe",
      water_source == 2 ~ "Outside private",
      water_source == 3 ~ "Public standpipe",
      water_source == 4 ~ "Well",
      water_source == 5 ~ "River, Lake, Spring, Pond",
      water_source == 6 ~ "Rainwater (Tank)",
      water_source == 7 ~ "Trucked water (NWC)",
      water_source == 8 ~ "Bottled Water",
      water_source == 9 ~ "Other (Specify)",

      TRUE ~ as.character(water_source) # keep original value if not matched
    ),

  )

print(head(SLC_2007.Subset))

```

# Data Visualization



### Description of Demographic Variables


```{r}

library(tidyverse)
library(gtsummary)
library(gapminder)

SLC_2007.Subset2 <- SLC_2007.Subset %>%
        rename("Area Code" = area_code)

table1 <- SLC_2007.Subset2 %>%
        select("Area Code") %>%
        filter(!(`Area Code` %in% c(4, 5))) %>% # exclude Area Codes 4 and 5
        tbl_summary(
                missing = "no"
        ) %>%
        add_n() %>%
        modify_header(label = "**Characteristic**") %>%
        bold_labels()

table1






```

```{r}
SLC_2007.Subset2 <- SLC_2007.Subset %>%
        rename("Type of Toilet" = type_toilet)
table2 <- SLC_2007.Subset2 %>%
        select("Type of Toilet") %>%
        tbl_summary(
                missing = "no"
        ) %>%
        add_n() %>% # add column with total number of non-missing observations
        modify_header(label = "**Characteristic**") %>% # update the column header
        bold_labels()



table2



```

```{r}
ggplot(SLC_2007.Subset, aes(x = kitchen_shared)) +
        geom_bar(aes(y = ..count..), fill = "#619CFF") +
        labs(
             x = "Kitchen Sharing",
             y = "Count Frequency") +
        geom_text(aes(label = ..count..), stat = "count", vjust = 0.01, colour = "black")+
        theme_minimal()




```
```{r}
library(ggplot2)

ggplot(SLC_2007.Subset, aes(x = toilet_shared)) +
        geom_bar(fill = "#619CFF") +
        labs(
             x = "Toilet Sharing",
             y = "Frequency") +
        scale_x_discrete(labels = c("Exclusive Use", "Shared")) +
        geom_text(aes(label = ..count..), stat = "count", vjust =0.01, colour = "black")+

        theme_minimal()




```



# Description of Key Variables





```{r}
# Calculate correlation coefficient and format as table
cor_table <- data.frame(Correlation = round(cor(SLC_2007.Subset$per_cap_con_all, SLC_2007.Subset$hh_size_all), 4))
rownames(cor_table) <- "per_cap_con_all vs. hh_size_all"
cor_table


```

```{r}
# Generate summary table of water_bill variable
summary(SLC_2007.Subset$water_bill)

```





```{r}
# Remove rows with missing values
SLC_2007.Subset <- na.omit(SLC_2007.Subset)

# Add first principal component
SLC_2007.Subset$pc <- predict(prcomp(~per_cap_con_all + water_bill, SLC_2007.Subset))[,1]

# Add density for each point
SLC_2007.Subset$density <- fields::interp.surface(
        MASS::kde2d(SLC_2007.Subset$per_cap_con_all, SLC_2007.Subset$water_bill), SLC_2007.Subset[,c("per_cap_con_all", "water_bill")])

# Plot with title
ggplot(SLC_2007.Subset, aes(per_cap_con_all, water_bill, color = pc, alpha = 1/density)) +
        geom_point(shape = 16, size = 5, show.legend = FALSE) +

        theme_minimal() +
        scale_color_gradient(low = "#2473c5", high = "#f2aeff") +
        scale_alpha(range = c(.25, .6)) +
        labs(#title = "Relationship between Per Capita Consumption and Water Bill with PC and density",
             x = "Per Capita Consumption",
             y = "Water Bill")





```




```{r}
# Remove rows with missing values
SLC_2007.Subset <- na.omit(SLC_2007.Subset)

# Add first principal component
SLC_2007.Subset$pc <- predict(prcomp(~hh_size_all + per_cap_con_all, SLC_2007.Subset))[,1]

# Add density for each point
SLC_2007.Subset$density <- fields::interp.surface(
        MASS::kde2d(SLC_2007.Subset$hh_size_all, SLC_2007.Subset$per_cap_con_all), SLC_2007.Subset[,c("hh_size_all", "per_cap_con_all")])

# Plot with title and updated y-axis label and variable
ggplot(SLC_2007.Subset, aes(hh_size_all, per_cap_con_all, color = pc, alpha = 1/density)) +
        geom_point(shape = 16, size = 5, show.legend = FALSE) +
        geom_smooth(fill='red') +
        theme_minimal() +
        scale_color_gradient(low = "#2473c5", high = "#f2aeff", guide = "none") +
        scale_alpha(range = c(.25, .6), guide = "none") +
        labs(#title = "Relationship between Household Size and Per Capita Consumption with PC and density",
                x = "Household Size",
                y = "Per Capita Consumption")

```
```{r}


# Remove rows with missing values
SLC_2007.Subset <- na.omit(SLC_2007.Subset)

# Add first principal component
SLC_2007.Subset$pc <- predict(prcomp(~hh_size_all + water_bill, SLC_2007.Subset))[,1]

# Add density for each point
SLC_2007.Subset$density <- fields::interp.surface(
        MASS::kde2d(SLC_2007.Subset$hh_size_all, SLC_2007.Subset$water_bill), SLC_2007.Subset[,c("hh_size_all", "water_bill")])

# Plot with title and updated y-axis label and variable
ggplot(SLC_2007.Subset, aes(hh_size_all, water_bill, color = pc, alpha = 1/density)) +
        geom_point(shape = 16, size = 5, show.legend = FALSE) +
        theme_minimal() +
        scale_color_gradient(low = "#2473c5", high = "#f2aeff") +
        scale_alpha(range = c(.25, .6)) +
        labs(title = "Relationship between Household Size and Water Bills with PC and density",
             x = "Household Size",
             y = "Water Bills")

```



```{r}
# Remove rows with missing values
SLC_2007.Subset <- na.omit(SLC_2007.Subset)

# Group water source categories
SLC_2007.Subset$water_source_group <- ifelse(SLC_2007.Subset$water_source %in% c("Public tap/standpipe", "Tube well/borehole"), "Piped water", SLC_2007.Subset$water_source)

# Transform water consumption variable by  logarithmic  transformation to compress the range of values.
SLC_2007.Subset$log_per_cap_con_all <- log(SLC_2007.Subset$per_cap_con_all)

# Create boxplot of water consumption by water source without outliers
ggplot(SLC_2007.Subset, aes(water_source_group, log_per_cap_con_all)) +
        geom_boxplot(fill = "#32aeff", alpha = 0.8, outlier.color = NA, outlier.shape = NA) +
        theme_minimal() +
        labs(title = "Distribution of Water Consumption by Water Source",
             x = "Water Source",
             y = "Log Per Capita Water Consumption")

```


```{r}
# Summarize water consumption by water source
water_summary <- aggregate(per_cap_con_all ~ water_source, SLC_2007.Subset, FUN = sum)

# Create bar graph of water consumption by water source
ggplot(water_summary, aes(x = water_source, y = per_cap_con_all, fill = per_cap_con_all)) +
        geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
        scale_fill_gradient(low = "#2473c5", high = "#f2aeff", guide = FALSE) +
        geom_text(aes(label = format(round(per_cap_con_all, 2), big.mark = ","), y = per_cap_con_all), vjust = -0.5, size = 3.5) +
        theme_minimal() +
        labs(#title = "Water Consumption by Water Source",
             x = "Water Source",
             y = "Total Water Consumption")


```

















