
---
output:
pdf_document: default
html_document: default
---


we c---
title: "Assignment 2.1"
output: html_notebook

---

Name: Sammarieo Brown
ID: 620142596
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```


### Package Management -> Importing the necessary packages that will be used in this project
```{r, warning = FALSE}
suppressWarnings(library(tibble))
suppressWarnings(library(tidyr))
suppressWarnings(library(haven))
suppressWarnings(library(readr))
suppressWarnings(library(dplyr))
suppressWarnings(library(ggplot2))
suppressWarnings(library(tidyverse))
suppressWarnings(library(ggrepel))
suppressWarnings(library(sjPlot))
suppressWarnings(library(sjmisc))
suppressWarnings(library(sjlabelled))
suppressWarnings(library(corrplot))
suppressWarnings(library(huxtable))
suppressWarnings(library(rstatix))
suppressWarnings(library(psych))
suppressWarnings(library(rempsyc))
suppressWarnings(library(kableExtra))
suppressWarnings(library(ggstatsplot))
suppressWarnings(library(ggpubr))
suppressWarnings(library(gtsummary))
suppressWarnings(library(see))
suppressWarnings(library(performance))
```

### Date Pre-processing & Preparation
1. Import dataset (SLC_2007.sav)
2. Rename column headers to be more descriptive.
3. convert the .sav file to a .csv file

```{r}
dataframe <- read_sav("F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.sav")

# rename column names to a more descriptive column name

colnames(dataframe)[1:45] <- c("HH_Num", "Final_Weight", "Water_Bill", "HH_Size_All", "HH_Size_Mem", "Per_Cap_Con_All", "Per_Cap_Con_Mem", "Per_Cap_Pop_Quint", "Type_Dwelling", "Material_Walls", "Num_Rooms", "Type_Toilet", "Toilet_Shared", "Kitchen_Shared", "Own_Dwelling", "Other_Dwelling", "Renter", "Rent_Amt", "Rent_Period", "Rent_Helper", "Pay_Mortgage", "Mortgage_Payment", "Mortgage_Num", "Mortgage_Period", "Pay_Taxes", "Tax_Period", "Water_Source", "Water_Lock", "Water_Meter", "Water_Bill_Latest", "Water_Bill_Months", "Water_Source_Shared", "Water_Source_Dist", "Water_Source_Dist_Code", "Light_Source", "Electric_Bill", "Light_Bill_Months", "Own_Land_Tele", "Own_Cell_Tele", "Land_Tele_Bill", "Cell_Tele_Bill", "Land_Tele_Use", "Cell_Tele_Use", "Garbage_Dispose", "Area")

# convert all the column names to lower case
colnames(dataframe) <- tolower(colnames(dataframe))

# coverting the SLC_2007.sav to csv

converted_df <- write.table(x=dataframe,file="F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.csv", quote = FALSE, sep = ",")

```
```{r}
# Set the seed and take 90% of the random sample

set.seed(710)

# import dataset
SLC_2007.Data <- read.csv("F:/DataSpell/Statistical Computing Project/dataset/SLC_2007.csv")


# random sample
n <- nrow(SLC_2007.Data)
sample_size <- round(0.9 * n) # calculate the desired sample size as 90% of the total number of rows

SLC_2007.Sample <- SLC_2007.Data[sample(seq_len(n), size = sample_size, replace = FALSE),] # take a random sample of the data set without replacement



```
# Subset of variables
```{r}
# create a sub dataset with the demographic variables
SLC_2007.Subset <- select(SLC_2007.Sample, type_dwelling,
                         toilet_shared,kitchen_shared,
                          water_source,hh_size_all,
                          per_cap_con_all,water_bill,water_source,area
)


```

# Inferential Analysis

### Goal 1: To determine if there is a difference in per capita water consumption based on location (area).
- #### Test: One-way ANOVA

```{r}

# Select the columns from your dataset
SLC_2007.Goal_1 <- select(SLC_2007.Subset, area, per_cap_con_all)

SLC_2007.Goal_1 <- SLC_2007.Goal_1 %>%
  mutate(area = recode(area,
                               `1` = "KMA",
                               `2` = "Other Town",
                               `3` = "Rural",
                       .default = "Unknown" # Add a default value for any unmatched value

  )
  )
SLC_2007.Goal_1  <- SLC_2007.Goal_1 %>%
  filter(area != "Unknown")


# chekc for normality
# Run the Linear Model
aov.model.test <- lm(per_cap_con_all ~ area, data = SLC_2007.Goal_1)


# Normality by Groups

# Check normality by groups, ignoring NA values
ggqqplot(na.omit(SLC_2007.Goal_1 ), "per_cap_con_all", facet.by = "area")

```
```{r}
# chec for equal variance

plot(aov.model.test,1)
```
```{r}
# One-Way ANOVA Test
aov.test <- aov(per_cap_con_all ~ area, data = SLC_2007.Goal_1)
summary(aov.test)

# Calculate r^2
RSq <- var(predict(aov.test)) / var(SLC_2007.Goal_1$per_cap_con_all, na.rm = TRUE)
round(RSq, 4)
```
```{r}
# If significant, run post hoc (multiple comparisons) test
pairwise.t.test(SLC_2007.Goal_1$per_cap_con_all, SLC_2007.Goal_1$area, p.adjust.method = "bonferroni")
```

```{r}
# Load ggplot2 and ggpubr
library(ggplot2)
library(ggpubr)

# Create the plot using ggbetweenstats()
anova_plot <- ggbetweenstats(
  data = SLC_2007.Goal_1,
  x = area,
  y = per_cap_con_all,
  type = "parametric",
  var.equal = TRUE,
  plot.type = "box",
  pairwise.comparisons = TRUE,
  p.adjust.method = "bonferroni",
  pairwise.display = "significant",
  centrality.plotting = FALSE,
  bf.message = FALSE
)

# Modify y-axis title
anova_plot <- anova_plot +
  ylab("Per Capita Water Consumption") +
  xlab("Area")

# Customize the theme to enlarge elements
anova_plot <- anova_plot +
  theme(
    text = element_text(size = 16), # Increase base text size
    axis.title = element_text(size = 18), # Increase axis title size
    axis.text = element_text(size = 14), # Increase axis text size
    plot.title = element_text(size = 20, face = "bold"), # Increase plot title size
    strip.text = element_text(size = 16), # Increase facet label text size
    legend.text = element_text(size = 14), # Increase legend text size
    legend.title = element_text(size = 16), # Increase legend title size
    panel.spacing = unit(1, "lines") # Increase space between facets
  )

# Display the plot
anova_plot

```



### Goal 2: To determine if there is a difference in per capita water consumption based on whether toilet facilities are shared or not.
- #### Test:  Independent Sample t-test

```{r}


# Prepare the dataset
SLC_2007.Goal_2 <- select(SLC_2007.Subset, toilet_shared, per_cap_con_all)
SLC_2007.Goal_2 <- SLC_2007.Goal_2 %>%
        mutate(toilet_shared = recode(toilet_shared,
                                      `1` = "EXCLUSIVE USE",
                                      `2` = "SHARED",
                                      .default = "Unknown" # Add a default value for any unmatched value
        )
        )

# Run the independent sample t-test
t_test_result <- SLC_2007.Goal_2 %>%
        filter(toilet_shared != "Unknown") %>% # Remove rows with "Unknown" values
        tbl_summary(
                by = toilet_shared,
                type = c(per_cap_con_all = "continuous"),
                statistic = list(per_cap_con_all ~ "{mean} ({sd})"),
                missing = "no",
                label = list(
                        per_cap_con_all ~ "Per Capita Water Consumption"
                )
        ) %>%
        add_difference()

# Display the result
t_test_result




```
```{r}
# Run the independent sample t-test using Version 2
t_test_result_v2 <- nice_t_test(data = SLC_2007.Goal_2,
                                response = "per_cap_con_all",
                                group = "toilet_shared",
                                warning = FALSE) %>%
  nice_table()

# Display the result
t_test_result_v2
```



### Goal 3:To determine if there is a relationship between per capita water consumption and household size (add control – area, toilet, kitchen)
- #### Test:  Multiple Linear Regression



```{r}
SLC_2007.Goal_3 <- select(SLC_2007.Subset, per_cap_con_all,hh_size_all,area,toilet_shared,kitchen_shared)
# recode variables.
SLC_2007.Goal_3 <- SLC_2007.Goal_3 %>%
  mutate(
    toilet_shared = recode(toilet_shared,
                                `1` = "EXCLUSIVE USE",
                                `2` = "SHARED"
  ),
    kitchen_shared = recode(kitchen_shared,
                                 `1` = "EXCLUSIVE USE",
                                 `2` = "SHARED",
                                 `3` = "NONE"
    ),
    area = recode(area,
                     `1` = "KMA",
                     `2` = "Other Town",
                     `3` = "Rural",
                     .default = "Unknown" # Add a default value for any unmatched value

    ),
    hh_size_all = case_when(
           hh_size_all == 1 ~ "1 person",
           hh_size_all == 2 ~ "2 person",
           hh_size_all == 3 ~ "3 person",
           hh_size_all >= 4 ~ "4 or more"
    )
  )


SLC_2007.Goal_3  <- SLC_2007.Goal_3 %>%
  filter(area != "Unknown")

print(head(SLC_2007.Goal_3))

```
```{r}
# Convert nominal variables to factors
SLC_2007.Goal_3_recoded <- SLC_2007.Goal_3 %>%
  mutate(
    toilet_shared = as.factor(toilet_shared),
    kitchen_shared = as.factor(kitchen_shared),
    area = as.factor(area),
    hh_size_all = as.factor(hh_size_all)

  )

# Run the multiple linear regression model
model <- lm(per_cap_con_all ~ hh_size_all + area + toilet_shared + kitchen_shared, data = SLC_2007.Goal_3)
tab_model(model)

```
```{r}
# Diagnostic Plots

check_model(model)
```


